### 2023-09-01 manuscript.tex
Expanded discussion of using Mirage2 per reviewer comment

The most popular aligners for protein sequences (Clustal Omega~\cite{Sievers2017}, MAFFT~\cite{Katoh2013}, MUSCLE~\cite{Edgar2004}, T-Coffee~\cite{Notredame2000}) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons. Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage}~\cite{Nord2018}. The biggest challenge in practice, however, is mapping a protein sequence to its genomic sequence to identify splice sites in the protein sequence. Although this information can in principle be derived from the GTF annotation files produced by the NCBI pipeline, annotating the splice junctions in the protein sequences themselves would facilitate splice-sensitive alignment.

>>>>

The most popular aligners for protein sequences (Clustal Omega~\cite{Sievers2017}, MAFFT~\cite{Katoh2013}, MUSCLE~\cite{Edgar2004}, T-Coffee~\cite{Notredame2000}) do not include splice sites in their alignments, which makes them prone to aligning non-homologous exons. Current algorithms can easily be extended to incorporate splice sites by coding them as a new symbol and preventing alignment between splice sites and amino acids, which was recently implemented in the aligner, \textit{Mirage2}~\cite{Nord2018, Nord2023}. The biggest challenge in practice, however, is mapping a protein sequence to its genomic sequence to identify splice sites in the protein sequence. Although this information can in principle be derived from the GTF annotation files produced by the NCBI pipeline, annotating the splice junctions in the protein sequences themselves would facilitate splice-sensitive alignment. Though Mirage2 contains utilities which perform this mapping automatically, the NCBI annotation pipeline sometimes modifies the coding sequences of transcripts to correct for possible protein-altering mismatches or indels, labeling the resulting sequences as ``low quality proteins.'' The impact of these corrections on Mirage2's mapping algorithms is unclear, and as Mirage2 was released while this manuscript was in preparation, we were unable to fully investigate its performance in such cases. However, we strongly recommend that future studies aligning orthologous protein with variable exon structures explore the benefits of using Mirage2 relative to the MAFFT-based approaches described here.
